Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-12 paper/paper.tex
Killed code and most of implementation section. Started elaborating on compound data.

\title{Optimizing Abstract Abstract Machines}

@@

\paragraph{Primitives}

\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\begin{align*}
\sNum &\in \hat\interpdelta(\saddone,\mnum) &
\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
\strue &\in \hat\interpdelta(\szerohuh,\szero) &
\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

%% \begin{figure}
%% \begin{alltt}
%%   ;; State \(\to\) \(\Set(\)State\()\)
%%   (define (step \(\mstate\))
%%     (match \(\mstate\)
%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
%%        (match \(\mexp\)
%%          [(var\(\superscript\mlab\) \(\mvar\))
%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
%%       [(co \(\msto\) \(\mkont\) \(\mval\))
%%        (match \(\mkont\)
%%          ['mt (set (ans \(\msto\) \(\mval\)))]
%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
%%          [(fn\(\superscript\mlab\) \(\mval'\))
%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
%%          [(fi\(\superscript\mlab\) c a \(\rho\))
%%           (for/set ((k (get-cont \(\msto\) l)))
%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
%%        (match fun
%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
%%          [(op \(\mop\))
%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
%%          [_ (set)]))]))
%% \end{alltt}
%% \caption{Implementation of machine transition relation.}
%% \end{figure}

\newpage
\section{Baseline analyzer}
\label{sec:baseline}

@@

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. Let $wn :
\widehat{State} \times Store \to State$ be the metafunction that adds
a store to a storeless state, $nw$ removes the store, and $c.\msto$
projects the store out of a state $c$. Formally,

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
\end{align*}

@@

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
either ANF or CPS.

@@

\section{Precision preserving recipe}
\label{sec:opt}

@@

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
\end{itemize}
Each technique requires minimal changes to the reduction
relation/fixpoint, and each improves performance by integer
factors. The first three remain in the purely functional realm. The
final three use mutation in small, localized places, so they do not
increase conceptual complexity drastically.

@@

The first problem that we realized with the AAM approach was that much
of the non-determinism is spurious. For example, in a function
application,
\begin{center}{\tt (f x y)}\end{center}

where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
only to be joined back together in the application position. We would see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

For this, we introduce a new value for a delayed lookup that acts to
quotient the state space by the values at the delayed address.

@@

% Lazy:  cpu time: 32481 real time: 32881 gc time: 547

@@

\newpage

@@

\subsection{Pre-allocating the store}
Hashes are not the most efficent for raw speed. If we commit to using
sequential natural numbers for variables and expression labels, by the
time we finish parsing the input program, we will know exactly how
many addresses we will need. Thus the store can be represented by a
vector the size of the largest label assigned, plus 1. This technique
only works in 0-CFA, since there are many more addresses that can be
allocated in 1-CFA and higher. To support general $k$-CFA in a vector
representation, we need a growable vector and a map from addresses to
assigned indices. When assignable indices run out, we grow the
vector. The cost of this map is less than a hash representation for
the whole store since typically there are more references to an address
than allocations of it.

% Ian's sections: generators, implementation and compound data
\section{Implementation}
\label{sec:impl}

All the above techniques can be combined into a single implementation
of the small step reduction relation and simple variations of a
general fixpoint combinator. We implemented all of these ideas in an
analysis of core Scheme (call-by-value lambda calculus extended with
variable-arity functions, call/cc, vectors, conses, recursive binding,
various primitives) in just 3000 lines of Racket. Most lines are spent
defining the primitives and expected library functions used in our
benchmarks. Here we describe the unifying vocabulary for writing such
a framework, and some design decisions related to scaling to a
language with compound data.

\subsection{Vocabulary of Parameterized Optimizations}

We combine the interface of generators~\cite{ianjohnson:cluhistory}
with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
of {\tt ev} states. We also have a special {\tt do} form that performs
behind-the-scenes state passing and iteration much like {\tt do} in
Haskell. In the presence of the different optimization
techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
change meaning slightly, but keep the same interface as their names
suggest.

\begin{figure}
\begin{alltt}
\(e-dispatch =\)
(match e
  [(var\(\superscript\mlab\) x)
   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
    (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
      (yield (co \(\msto\) k v))))]
  [(lit\(\superscript\mlab\) l)
   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
    (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
  [(lam\(\superscript\mlab\) x e)
   (define c (compile e))
   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
    (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
  [(app\(\superscript\mlab\) f e)
   (define cf (compile f))
   (define ce (compile e))
   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
    (define-values (\(\msto\)* a) (push state))
    (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
  [(ife\(\superscript\mlab\) e0 e1 e2)
   (define c0 (compile e0))
   (define c1 (compile e1))
   (define c2 (compile e2))
   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
     (do (\(\msto\)) [(\(\msto\)* a) (push state)]
      (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])

(define (step state)
  (match state
    [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
    [(co \(\msto\) k v)
     (match k
       ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
        (yield (ans \(\msto\) v))))]
       [(ar\(\superscript\mcntr\subscript\mlab\) e \(\menv\) a)
        (generator
         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{f}\) (force \(\msto\) v))])
           (yield (ev \(\msto*\) e \(\menv\) (fn\(\superscript\mcntr\subscript\mlab\) a\(\superscript{f}\) a)))))
       [(fn\(\superscript\mcntr\subscript\mlab\) fa a)
        (generator
         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{a}\) (force \(\msto\) v))]
                  [k <- (get \(\msto\) a)]
                  [f <- (get \(\msto\) fa)])
           (yield (ap\(\superscript\mcntr\subscript\mlab\) \(\msto*\) f a\(\superscript{a}\) k))))]
       [(fi\(\superscript\mcntr\subscript\mlab\) c a \(\menv\))
        (generator
         (do (\(\msto\)) ([k <- (get \(\msto\) l)]
                  [v <- (force \(\msto\) v)])
           (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
    [(ap\(\superscript\mcntr\subscript\mlab\) \(\msto\) fun a k)
     (match fun
       [(clos x e \(\menv\))
        (generator
         (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
           (yield (ev \(\msto\)* e \(\menv\)* k))))]
       [(? op? o)
        (do (\(\msto\)) ([k <- (get \(\msto\) l)]
                 [v <- (force \(\msto\) v)]
                 [v (\(\interpdelta\) o (list v))])
          (yield (co \(\msto\) k v)))
       [_ (generator (do (\(\msto\)) () (continue)))]))]))
\end{alltt}
\caption{Parameterized Abstract$^2$ machine for ISWIM}
\label{fig:paam}
\end{figure}

\begin{figure}
\begin{alltt}
(define (compile e) \(e-dispatch\))
(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\machstep\) (\(\lambda\) (\(\msto\) \(\menv\) k) body \ldots)
(yield (ev \(\msto\) c . rest)) \(\longmapsto\) (c \(\msto\) . rest)
(yield e) \(\longmapsto\) (yield-meaning e)
\end{alltt}
\caption{Meaning of forms if compiled}
\label{fig:cfm}
\end{figure}

\begin{figure}
\begin{alltt}
(define (compile e) e)
(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\longmapsto\) (let () body ...)
(yield e) \(\longmapsto\) (yield-meaning e)
\end{alltt}
\caption{Meaning of forms if not compiled}
\label{fig:ncfm}
\end{figure}

Why generators? Relations in general cannot be written as computable
functions, but in our case we have a decidable relation with finitely
many pairs sharing the same left component. There are several
implementation strategies for such relations, some more effective than
others. For instance, a silly implementation strategy would be
defining a binary function that decides membership in the relation. We
want the definition to drive the interpreter, so we might naturally
decide to write a function that takes a state and returns a set of all
the next states. As we described in the imperative workset section,
this intermediate allocation is unnecessary and wasteful. What we
really want is a way of pulling on the reduction relation for more
next states as the fixpoint is ready for them.

For intermediate sets, {\tt yield} is {\tt return} in the set
monad. For the imperative worklist, {\tt yield} performs mutation
behind the scenes. Indeed, we could use real generators! This is the
only case where {\tt generator} constructs a nullary generator; in the
other cases it evaluates to its body. Racket's generators are not
tuned for performance, so we actually lose efficiency in this case,
but the unifying vocabulary is instructive nonetheless.

The {\tt $\lambda$\%} form constructs a real closure in the
compilation case and evaluates to its body otherwise.

The {\tt do} form has the special job of maintaining the current store
and accumulated states (if indeed those are being passed around). It
uses Racket's syntax parameters ~\cite{dvanhorn:eli/stxparam} to
track the most recent bindings of these accumulators so that {\tt
  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
  statep}). In the case of state passing and set-accumulated states, its
definition is simply

\begin{center}
\begin{alltt}
(yield-meaning e) \(\longmapsto\) (values \(\sigma\)p (set-add statep e))
\end{alltt}
\end{center}

The linguistic elements described above were all reusable to implement
the various languages' abstract interpreters. The language of our
benchmarks, Scheme, has compound data (e.g. vectors and conses), which
introduces choices for abstraction.

@@

We discovered that the choice of abstract domain for literal data
(such as the large quoted axiom database in the Boyer benchmark)
drastically affected the performance of our analysis on the benchmark
suite. It is worth discussing the ``obvious'' and na\"ive abstraction
that is natural to a reader of
~\citep{dvanhorn:VanHorn2011Abstracting}, and other easy-to-implement
yet effective approaches.

@@

The uniform way they approach a simple abstraction strategy is to cut
recursion out of the data definition by tying the recursive knot
through the abstract store. For Scheme, the grammar for values looks like the following:

@@

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic
callsites, and quoted lists really are sugar for a sequence of calls
to {\tt cons}, this abstraction explodes the value space. Analyzing a
function that counts the number of atoms in a literal s-expression
would actually interpret that function at least that number of times
(more because of intermediate conses). Indeed, even in our fastest
parameterization, this abstraction causes the analysis of Boyer to be
430 times slower than the approach we will now describe.

@@

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is dramatically smaller in typical Scheme programs. We
use the above abstraction for these syntactic uses, but choose a
different strategy for quoted lists. Quotation is special because it
cannot introduce function values or mutable values, which is important
to enhancing our technique's soundness to conceptual complexity ratio.

@@

The alternative to {\tt QData} is to go higher in the lattice to
$\top$, which has more complicated meaning (Shivers' escape semantics) and
is overly approximate.

@@

% \include{appendix}

>>>>

\title{Optimizing Abstract$^{\text 2}$ Machines}

@@

%% \paragraph{Primitives}

%% \begin{align*}
%% \mnum+1 &\in \interpdelta(\saddone,\mnum) &
%% \mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
%% \strue &\in \interpdelta(\szerohuh,\szero) &
%% \sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
%% \end{align*}

%% \begin{align*}
%% \sNum &\in \hat\interpdelta(\saddone,\mnum) &
%% \sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
%% \strue &\in \hat\interpdelta(\szerohuh,\sNum) &
%% \sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
%% \strue &\in \hat\interpdelta(\szerohuh,\szero) &
%% \sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
%% \end{align*}

%\newpage
\section{Baseline analyzer}
\label{sec:baseline}

@@

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. The
scaffolded reduction relation, $\widehat{\machstep}$, is as follows:

Let $wn : \widehat{State} \times Store \to State$ be the metafunction
that adds a store to a storeless state and $c.\msto$ projects the
store out of a state $c$.

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

@@

The important rules that change are the following, specialized to 0-CFA:

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
either ANF or CPS. We can also play a representation trick and remove
$\menv$ since we know it is the identity function in 0-CFA. We will
refer to this change in the semantics as value indirection.

@@

\section{Mostly precision preserving recipe}
\label{sec:opt}

@@

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
\end{itemize}

Each technique requires minimal changes to the reduction
relation/fixpoint, and each improves performance by integer
factors. The first three remain in the purely functional realm. The
final three use mutation in small, localized places, so they do not
increase conceptual complexity drastically.

@@

A shortcoming of the AAM approach is that much of the non-determinism
is spurious. For example, in a function application, {\tt (f x y)},
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
only to be joined back together in their respective application positions. We would
see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

For this, we introduce a new value for a delayed lookup that acts to
quotient the state space by the values at the delayed address. Its changes to the semantics are highlighted in the following rules:

@@

This semantics introduces a subtle precision difference over the
baseline that is negligible. Consider a configuration where a
reference to a variable and a binding of a variable will happen in one
step. With laziness, the reference will mean the original binding(s)
of the variable or the new one, because the actual store lookup is
delayed precisely one step (i.e. laziness is administrative). Without
laziness, the reference will fan out to all the bindings of the
variable before the new binding happens and thus might have an
observable precision difference. In our experience, this situation
happens too rarely to warrant the significant increase in time and
memory needed for this eager non-determinism. Indeed, were the
variable reference a step later and another binding is not made in
that step, the results of the two approaches are the same.

The administrative nature of laziness means that we could remove the
loss in precision by essentially duplicating the reduction relation to
specialize variable lookup. This works since in the semantics of ISWIM
with value indirection consumes the value component of states in
one step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

All following techniques preserve precision.
% Lazy:  cpu time: 32481 real time: 32881 gc time: 547

@@

%\newpage

@@

\subsection{Pre-allocating the store}
Hashes are not the most efficent for raw speed. If we commit to using
sequential natural numbers for variables and expression labels, by the
time we finish parsing the input program, we will know exactly how
many addresses we will need. Thus the store can be represented by a
vector the size of the largest label (call it $\mlab_{max}$) assigned, plus
1. The parser should be assigning new, fresh names to variables
anyway, so this change is a one-liner.

This technique by itself only works in 0-CFA, since there are many more
addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
many combinations of labels are not likely to occur. The all up-front
allocation again breaks down when we add compound data, which
introduce new kinds of labels (e.g. left and right components of a
tuple).

@@

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
value of which do not particularly matter for control-flow
analysis. If we interpret a list literal of 30 elements (that again
have 10 elements each) as 300 applications of {\tt cons}, and then
interpret a function that traverses the tree, we will end up
interpreting that function at least 300 times, when once or twice
would do.

We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k-CFA$
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

@@

The uniform way AAM approaches a simple abstraction strategy is to cut
recursion out of the data definition by tying the recursive knot
through the abstract store. For Scheme, the grammar for values looks
like the following:

@@

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic callsites
in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
for a sequence of calls to {\tt cons}, this abstraction explodes the
value space. Analyzing a function that counts the number of atoms in a
literal s-expression would actually interpret that function at least
that number of times (more because of intermediate conses). Indeed,
even in our fastest parameterization, this abstraction causes the
analysis of Boyer to be 430 times slower than the approach we will now
describe.

@@

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is dramatically smaller in typical Scheme programs. We
use the above abstraction for these syntactic uses, but choose to
interpret literal lists as not always sugar for cascading {\tt
  cons}es. In particular, if a list literal is ``too big'' (in our
case length > 1), we interpret the list as a circular data structure;
the right address points back to the {\tt cons} itself, and the left
address points to all of the elements of the list. We take this
farther and join the list elements together in with a coarse
type-based abstraction. Quotation is special because it cannot
introduce function values, which is important to enhancing our
technique's soundness to conceptual complexity ratio. If we join two
values of different type together, we don't get ``anything,'' which
has complex meaning in higher-order languages (Shivers' escape
semantics) and is overly approximate. We instead get ``any quotable
value,'' which has much simpler semantics.

@@

@@

\appendix
\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
\label{sec:accept}

\cite{dvanhorn:nielson-nielson-popl97} \cite{dvanhorn:Neilson:1999}

This machine's allocation strategy mimics the Uniform k-CFA analysis
in Principles of Program Analysis, which is defined in terms of
``$\delta$ contours.''  However, because the machine represetation makes
context explicit via continuations, we can calculate these contours
rather than thread them throught the evaluator.  In other words, we
can use the CESK* machine without modification to obtain Uniform k-CFA
by way of a simple allocation strategy.  (In this way, it's a
simplification of the presentation in JFP.)

NNH uses a coinductive acceptability relation to specify Uniform
k-CFA:

\[
   C,R \models^{ce}_\delta E
\]

The cache and global environment form a finite store-like structure
holding bindings and return values.  The contour environment ce maps
variables to locations in R which contains their bindings, just as the
environment of the CESK* machine does.  The current contour delta is a
string of application labels describing the enclosing context under
which this term is being analyzed (or evaluated).  If you view the
acceptability relation as a big-step evalator, the
$(\widehat C,\widehat\rho)$ component should be seen as a global
store ce is the environment mapping variables to their locations.

Starting form the initial configuration for a program and iterating
the machine transition relation until reaching a fixpoint of reachable
states will \emph{underestimate} the acceptability relation of Uniform
k-CFA.  You can recover acceptability by feeding this store back into
the initial configuration and iterating again.  Repeating this process
until a complete run of the program reaches no new states will be the
least solution that is acceptable.

HOWEVER.  Why should we care about acceptability?  What this
machine computes is safe.  In other words, it computes a more
precise characterization of the run-time behavior of a program.  In
doing is so, it actually saves work (as can be seen above).

An Example:

\begin{alltt}
 (let ((id (\(\lambda\) (x) x)))
   (begin (id 1) (id 2)))
\end{alltt}

Under Uniform 0-CFA, we would have:
\[
   [{\tt x} \mapsto \{{\tt 1}, {\tt 2}\}] \in \widehat\rho
\]

in the least solution to $\models$.  This says that, when run, 'x' is
bound to 1 or 2.

Under the machine semantics using a 0CFA allocation policy, the trace
semantics of the machine show that x is bound to 1, and that at some
later point, x becomes bound to 1 or 2.  Moreover, the machine would
show that (id 1) evaluates to 1 and only 1, while Uniform 0CFA must
give that (id 1) is either 1 or 2 to be acceptable.  We don't see any
value in these kinds of false flows that are due to the global and
timeless aspects of C,R which acceptability requires the heap to be
both finite and unchanging over the course of abstract
interpretation. (Another view of the difference: the machine abstracts
a program's execution as a \emph{finite state machine} that mimics the
machine interpretation of the program; the aceptability relation of
Uniform \(k\)-CFA abstracts a program's execution as a \emph{finite
  map} that mimics the big-step evaluator: from terms to (sets of)
values.)

\subsection{Another problem with acceptability: Temporal ignorance}

The small-step approach to static analysis brings subtle yet important temporal
richness not found in classical analyses for higher-order programs.
%
Classical analyses (ultimately) compute judgments on program terms and
contexts, e.g., at expression $e$, value $x$ may have value $v$.
%
The judgments do not relate the order in which expressions and context may be
evaluated in a program, e.g., a classical analysis has nothing to say with
regard to question like, ``Do we always evaluate $e_1$ before $e_2$?'' or ``Is
it always the case that a file handle is opened, read and then closed in that
order?''

Small-step analyses, by their nature, encode the temporal relationships between
abstract states.
%
It is sensible to make temporal queries of a small-step analysis.
%
Of course, this does not come for free: respecting temporal order imposes an
order in which states and terms may be evaluated \emph{during} the analysis.
%
Classical analyses can (and do) evaluate expressions in any order, or in some
cases, even in parallel~\cite{might:Prabhu:2010:EigenCFA}.
%
Relaxing that restriction on order affords additional optimizations that we
have \emph{not} performed.

We avoid sacrificing order not simply because we are interested in the
questions it allows us to ask, but because considering temporal order actually
improves the precision of the analysis itself.

\section{Pushdown Analysis}

It is straightforward to instantiate a \emph{pushdown} abstraction by
bounding only the variable binding portion of the heap, but using a
unique allocation strategy for continuations.  Such a strategy
abstracts a program's execution as a \emph{pushdown automata}
that mimics the machine interpretation.  This strategy therefore
models the abstract stack in a true stack like fashion and always
properly matches function calls with their return.

Although such analyses can be formulated straightforwardly in the
abstract machine approach, it is not clear all of the techniques of
this paper can be applied to similar effect in the pushdown context.
The main problem is calculation of an analysis can no longer be
computed as the fixed point of the machine transition relation.
Although there are several implementations (CFA2,ICFP'12), they
operate at speeds roughly on par with our starting point: unoptimized
store widened
machines. \cite{dvanhorn:Earl2012Introspective,dvanhorn:Vardoulakis2011CFA2}
